EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-11-08 BrainScience/response.letter.3.tex

2) The write-up on the active modality section is currently slightly vague. For instance, the reader should not be required to consult Riccio et al. 2013 to obtain information on what "feedback" constitutes. Also, perhaps I am missing something, but I am slightly confused as to why pseudo-real datasets involving superimposition of a selected P300 waveform template are being constructed here. If this is a replication of Riccio, doesn't the original stream itself contains naturally occurring P300s which the authors could contrast the detection performance of their selected classification methods of choice on?

3) As indicated in my initial summary section, the value of this work to the clinical community remains unclear. The value of BCI in clinical work has a long and well-established history (e.g., Neuper et al. 2003). The reference cited in support of this in the response document (Chavarriaga et al., 2017) focuses on the challenges of long-term independent BCI use by intended end users which the current study does not seem to focus on.

\subsection*{\ovalbox{General Comments}}
The authors have addressed many of my concerns regarding this manuscript. There are however a few points that could benefit from clarification. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We thank again the Reviewer for all the detailed information provided on this fruitful Review.

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
1) While the authors have addressed the issue of having just a single null-EEG stream from a single subject, it appears to be the case still that a single P300 ERP template is being used. This would seem to either assuming that the P300 is generally invariant across trials (which is unlikely) or that the result of this assessment is restricted to the case in which the ERP response of interest does not vary between trials. This should at least be discussed as a limitation in the conclusion section of the manuscript. 
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We have modified the experiments to use all the available templates.  On the passive-modality, we are using now 70 averaged templates obtained from Subject Number 8 of the BNCI-Horizon dataset (Riccio et al).  They are randomly selected and they are superimposed on the EEG stream.  This subject was selected due to the fact that the waveforms of their averaged P300 responses have a very distinctive and prototypical shape.

For the active-modality, first all 70 templates are extracted from each subject by an offline averaging procedure.  These 70 templates are superimposed to a random non-hit preprocessed segment (see next issue).

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2) The write-up on the active modality section is currently slightly vague. For instance, the reader should not be required to consult Riccio et al. 2013 to obtain information on what "feedback" constitutes. Also, perhaps I am missing something, but I am slightly confused as to why pseudo-real datasets involving superimposition of a selected P300 waveform template are being constructed here. If this is a replication of Riccio, doesn't the original stream itself contains naturally occurring P300s which the authors could contrast the detection performance of their selected classification methods of choice on?

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We have modified the Section 2.6 to clarify in detail the procedure.  We are using a pseudo-real dataset to have certainty of the existence of the P300 response, which is generating by a subject and we are not sure if the response is there or not.  This is the reason we followed a similar procedure performed by Varlaager in REF while trying to analyze different alternatives of alignment of P3 trials.  But instead of using a costumized completely artificial P300 waveform we used one obtained from a public dataset of a real person, and from a particular group which is targetted by BCI applications.

### 2018-11-08 BrainScience/brainscience.tex
More changes

\begin{itemize}
\item EEG Stream: the raw EEG streams for both modalities
\item Templates: P300 templates obtained either from the Subject Number 8 or from the same subject.
\item Pseudo-real dataset
\item null-signals
\end{itemize}

The objective of the following experiments is to assess the performance of the algorithms that aim to recognize the shape of the P300 waveform, obtained after averaging signal segments.  This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.  The experiments are performed by offline evaluation of the performance on a P300-Based BCI Speller application.

@@

The problem with any P300 experiment is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.  Therefore, P300 experiments are often mined with \textit{null-signals}.  These are EEG streams which are marked as hit but, because the subject was not particularly focused, or concentrated or so on, and the real signal is not generated.   This lack of certainty may be in detriment of analysis of algorithms and can be misleading or difficult to dealt with.   Similar experiments had address this same issue, particularly when benchmarking different algorithms.

@@

First for a passive modality, a real P300 ERP template, obtained from a public dataset, is superimposed into a distinct EEG stream.  This trace was experimentally obtained by a subject which was observing the flashing of the stimulus matrix during a P300 Speller procedure but did not engage in focusing on any letter in particular. Everything is there, except the P300 ERP component. Along the EEG stream, the markers information was used to localize the \textit{True} segments where the P300 should have been found, and those timing locations are used to superimpose the extracted ERP waveform.  By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential in accordance to similar procedures used in other works \citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.

In this case the template ERP is acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. Segments from the EEG signal containing the ERP are extracted for the trial number $2$, and they are point-to-point coherently averaged.  This P300 ERP can be seen in Figure  \ref{fig:erptemplate1}.  This template is selected due to its shape which more closely resemble the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.

>>>>

%
%\begin{itemize}
%\item EEG Stream: the raw EEG streams for both modalities
%\item Templates: P300 templates obtained either from the Subject Number 8 or from the same subject.
%\item Pseudo-real dataset
%\item null-signals
%\end{itemize}
%

The objective of the following experiments is to assess the performance of the algorithms that aim to recognize the shape of the P300 waveform, obtained after averaging signal segments.  This performance is evaluated by processing a pseudo-real dataset with two modalities where subtle alterations on the latency and amplitude of the P300 component are simulated in a controlled environment.  The experiments are performed by the offline evaluation of the performance on a P300-Based BCI Speller application.

@@

A problem with any P300 Speller is that the subjects that take part on the experiment are within the closed loop of the BCI system and the human is  not a static compliant entity that always performs what the experimenter asks for in a precise and consistent way~\citep{Chavarriaga2017}.  Therefore, P300 experiments are often mined with \textit{null-signals}.  These are EEG streams which are marked as having the signal component but, because the subject was not particularly focused, or concentrated, the real signal is not generated.   This lack of certainty may be in detriment of analysis of algorithms and can be misleading or difficult to deal with.   Similar experiments had addressed this same issue, particularly when benchmarking different algorithms~\citep{Ouyang2017,Jaskowski2000,QuianQuiroga2003}.

To solve this issue a pseudo-real dataset is generated under two different modalities.  A passive modality and an active modality.

@@

First for a passive modality, real P300 ERP templates obtained from a public dataset, are superimposed into a distinct EEG stream.  A set of template ERPs are acquired from the Subject Number $8$ of the public dataset 008-2014~\citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by IRCCS Fondazione Santa Lucia. On the other hand, the EEG traces where these templates are superimposed, are experimentally obtained by subjects which are observing the flashing of the stimulus matrix during a P300 Speller procedure but they do not engage in focusing on any letter in particular. Everything is there, except the P300 ERP component. Hence, along the EEG stream, the markers information is used to localize the \textit{True} segments where the P300 should be found, and those timing locations are used to superimpose the extracted ERP waveform.  By implementing this pseudo-real approach, it is possible to effectively control null-signals and to adjust the shape of this evoked potential.

A sample P300 ERP obtained from the trial number $2$ of Subject 8 can be seen in Figure  \ref{fig:erptemplate1}.  The templates obtained by this processing this Subject's EEG signals are selected due to their shapes more closely resembling the prototypical P300 waveform~\citep{Rao2013,Clerc2016}.  Segments are extracted from the original dataset, and they are point-to-point coherently averaged.
